\section{Introduction}
Programs crash. Static type systems serve to make the life of a programmer
easier by providing static compile-time guarantees of certain program behaviour,
guaranteeing that a function will map from a domain to a co-domain. A
dependently typed programming language gives a programmer even additional
expressive power, allowing types to be expressed relative to values given at
runtime. For example, consider a function replicate, that takes in any Peano
natural $n$ and returns a vect of $n$ Peano natural of value 1 to $n$. We can
express this in Idris, a dependently typed functional programming language, as
follows in Figure ~\ref{replicate_in_idris}. 

\begin{figure}[ht!]
  \caption{Implementation of Replicate in Idris}
  \label{replicate_in_idris}
  \begin{lstlisting}
    replic : (n : Nat) -> Vect n Nat
    replic n = replicateHelper n n where
      replicateHelper : (n : Nat) -> Nat -> Vect n Nat
      replicateHelper Z _ = Nil
      replicateHelper (S n) val = (val :: replicateHelper n val)
  \end{lstlisting}
\end{figure}

Thus, the co-domain of our function is dependent upon some value that is
provided at runtime. Calling \texttt{replic 5} will limit the co-domain of the
function to all Vects of length 5, etc. Thus, having written this type, a
programmer now has compile-time guarantees of extreme correctness that would not
be otherwise possible. 

While research in dependent types has traditionally been in the domains of
theoretical mathematics, where researchers use dependently typed theorem provers
to write mathematical formulisms, dependent types have applications outside of
such a theoretical domain. In my work, I seek to summarize the existing body of
work around practical applications of dependent types. 

\section{Additional Work}
This review is a  tour of a potential application of dependent types. In this
abstract, I have chosen a brief example, units of measurement, in order to
demonstrate their potency. However, in addition to units of measurement,
dependent types have far ranging applications. My review explores the following
additional examples: 
\begin{enumerate}
  \item Complicated pattern matching in the Cryptol DSL for cryptographic
  applications. \cite{power_of_pi}
  \item Generating file format parsers from a data description language.
  \cite{power_of_pi}
  \item A relational database and algebra. \cite{power_of_pi, eisenberg2016}
  \item An alternative to monadic transformations with an algebraic effects DSL
  \cite{algebraic}
  \item Programming distributed systems with F\* and dependent types.
  \cite{fstar_distributed_programming}
  \item A low-level domain-specific language demonstrating the applicability of
  dependent types to systems programming. \cite{idris_systems_programming}
\end{enumerate}

\section{Units of Measurement}
A practical application for dependent types from is to eliminate bugs that can
arise from improper unit conversions \cite{gundry2013}. Units of measurement are
already implemented in Microsoft's F\# Programming Language
(\cite{kennedy2009}). If numbers carry a type denoting their unit of measurement
with them, we can ensure at compile time that improper unit conversions are not
going to occur at runtime. These bugs can be catastrophic, as made evident by
NASA's loss of a \$125-million \textit{Mars Climate Orbiter} when ``spacecraft
engineers failed to convert from English to Metric units of measurement''
\cite{hotz1999}. 

\begin{figure}[ht!]
  \caption{Displays a unit error that would be caught at compile-time with units of measurement.}
  \label{unit_error}
  \begin{lstlisting}
    distanceTraveled : Quantity Inches
    distanceTraveled = inches 20

    distanceLeft : Quantity Metres
    distanceLeft = (metres 1000) - distanceTraveled
  \end{lstlisting}
\end{figure}

An implementation of units of measurement should have types that support
decidable equality by definition. Two typed variables can only be equal because
they have the same unit of measurement or derived unit of measurement and the
same value. This means that if coding style guidelines enforce that all numeric
values must be well-typed with units of measurement, there will be compile-time
guarantees that errors of conversion between units of measurement will not
occur. See Figure ~\ref{unit_error} for an example of a program that should
error. 

While units of measurement are implemented as a feature in the F\# language,
which is not dependently typed, a dependently typed programming language would
allow for a units of measurement system to be implemented \cite{gundry2013}.
Gundry invites us to consider a system for describing units in terms of a
constructor that allows us to both enumerate elementary units and also express
derived units in terms of one another. 

\begin{figure}[ht!]
  \label{idris_code}
  \caption{Basic SI unit declarations in adapted from Dependent Haskell to Idris \cite{gundry2013}}
  \begin{lstlisting}
    data Unit : Int -> Int -> Int -> Type
    
    Dimensionless : Type
    Dimensionless = Unit 0 0 0
    
    Metres : Type
    Metres = Unit 1 0 0
    
    Seconds : Type
    Seconds = Unit 0 1 0
    
    Kilograms : Type
    Kilograms = Unit 0 0 1
    
    data Quantity u = Q Double
    
    metres : Double -> Quantity Metres
    metres v = (Q v)
    
    seconds : Double -> Quantity Seconds
    seconds v = (Q v)
    
    kilograms : Double -> Quantity Kilograms
    kilograms v = (Q v)
    
    plus : Quantity u -> Quantity u -> Quantity u
    plus (Q x) (Q y) = Q (x + y)
  \end{lstlisting}
\end{figure}

For now, unit only supports three elementary units (metres, seconds, kilograms),
but one can imagine a full library implementing the entire SI Units system. Each
elementary unit is implemented as a single 1 in the call to the Unit constructor
with all entries as zero. Thus, we can express derived units in a call to the
Unit constructor where negative integers would represent elementary units
present in the denominator. 

We can define quantities as a type containing a \texttt{Unit} and an integer.
This then allows us to write simple constructors for the quantity type. We can
then define well-typed multiplication and addition operations giving us similar
guarantees to that which is given by units of measurement in F\#. 

As defined above, this enforces well-typed addition, requiring that two
additions be of the same type. we can also define operations that allow us to
express fractional units. For example, a Newton of force is defined as a
$kg\times ms^{-2}$. Therefore, if we are able to compose types through
multiplication and division, we can express a Newton with our units system. See
Figure ~\ref{division}. 

\begin{figure}[h]
  \caption{Definition of division and multiplication of dependently typed units
  of measurement. Ported to Idris from \cite{gundry2013}}
  \label{division}
  \begin{lstlisting}
    Newtons : Type
    Newtons = Unit 1 -2 1

    newtons : Double -> Quantity Newtons
    newtons val = over 
        (times (kilograms val) (metres 1)) 
        (times (seconds 1) (seconds 1))
  \end{lstlisting}
\end{figure}

What we've shown here is that while units of measurement can be first-class
features in a programming language like F\#, a dependently typed language allows
us to build certain functionality easily into the language without changing the
language specification whatsoever. 

\section{Conclusions}
Through a review of the practical applications of dependent types in existing
and implemented functional programming languages, I demonstrate that while
dependent types are often thought of as far-off and theoretical, they are
accessible to the modern programmer. In addition to empowering constructive
mathematicians with theorem provers, dependent types give a programmer ways to
write extremely safe programs with certain compile-time guarantees of
correctness. 
