
\cleardoublepage

\addcontentsline{toc}{chapter}{Acknowledgements}

\begin{acknowledgements}

I am indebted to so many people who have done so much prior work in
the field I find myself in and to the plethora of people who've mentored me over
the past year. 

I am incredibly grateful to have had the experience of presenting the literature
review portion of this thesis at the Principles of Programming Languages Student
Research Competition (POPL SRC) in Los Angeles this past January. I was also
able to attend the co-located Programming Languages Mentoring Workshop through
the generous scholarships offered by the workshop. At PLMW, I had the
opportunity to chat with Nadia Polikarpova, Armando Solar-Lezama, and K. Rustan
M. Leino over Thai food. When they heard about my fascination with formal
verification and election systems, they were so encouraging and welcoming. The
programming languages community is truly a special place to be. I found myself
surrounded by people who were not only brilliant, but also incredibly
supportive. Their encouragement gave me even more fortitude that led to the
conclusion of this thesis, and for that, I sincerely offer them my gratitude.

I'd also like to offer my thanks to my POPL roommate, Yao Li. Through our
conversations, Yao helped me overcome impostor syndrome at POPL and really
helped me through my time at the conference. Yao is incredible and I'm excited
to read more of his work as he progresses at Penn. I also have to thank Rachel,
Xinyue and Divesh, who attended POPL with me and have been endless sources of
encouragement. 

When I first visited Haverford 4 years ago, I met Dave Wonnacott, who introduced
me to the Haverford Computer Science Department. I was struck by how accessible
he felt as a professor and how down-to-earth and kind he was. Dave is actually a
big part of why I chose to come to Haverford. Since coming to Haverford, I have
found a home in Haverford's small and fledgling Computer Science department. To
the other faculty who have taught and mentored me, Sorelle, Deepak, Liz, Jane,
Jean, Siddharth, JD, thank you. A special thanks to Sorelle, who after the
stunning and unexpected November 2016 election, discussed with me what it meant
as someone with skills in software engineering and computer science to use my
skills in the civic space. That very conversation had a great influence on my
subsequent passions quite possibly led to this thesis. Last, but not least,
Steven Lindell has run a fantastic Senior Seminar and has offered support and
guidance throughout the year. I also must thank the other thesising seniors for
being a source of support: Becky, Skyler, Dylan, Linyi, Madeleine, Mary and
Noah. 

To my advisor, Richard Eisenberg, I have inexpressible gratitude. Richard
introduced me to advanced functional programming and sparked my interest in this
field. Throughout this whole year, he has been an endless source of support and
mentorship. When I fail to understand a concept, I had the confidence that I
could walk into our weekly meetings, express my utter confusion and work through
that with him. Richard has a relentless attention to detail and a sharp mind and
he is usually able to instantaneously offer improvements to an idea that I've
been pondering over for a whole week. For students looking for a thesis advisor
who will seriously optimize around your learning and growth and challenge you to
express your ideas with clarity, you cannot go wrong with Richard. Richard, from
the bottom of my heart, thank you. 

To the board and the community at the Pan-Asian Resource Center, my teammates on
the Men's Rugby Team, the people I sing with in Chaverim and so many other
community members who have offered me the support spaces to learn and grow,
thank you. A special thanks to Rebecca, who has been an endless source of
support and accountability even from Peru. Haverford has shaken the very
foundations of my worldview to the ground and rebuilt them for the better. While
I have too many people to thank for that, these people deserve a special
mention. 

Decades ago, two university students sat on the grass next to the airport on a
little green island ruled by a brutal martial law. They watched planes take off
and they dreamt of leaving the island and searching for a world where
free-spirited characters could live happily. Those people were my parents, Misha
Fang-Chih Chou and Kenny Ghen-Yue Liao, to whom I am struggling to fathom the
words to even begin to thank. They immigrated for no other reason than the
pursuit of happiness. My entire life, they have always optimized not around my
success, but around what makes me happy. They have taught me that to live a
fulfilling life is not only about the relentless pursuit of milestones, but
about being content at the end of the day. 

I'm looking back at the choices I've made. I think I'm happy. I think I'm
content. 

Kevin Jiah-Chih Liao\\
April 2018

\end{acknowledgements}
