
\chapter{Conclusion}

\section{Conclusions on the Literature Review of Dependent Types}

While many literature reviews begin by looking examining a problem and looking
for existing solutions, this literature review takes an opposite approach. The
broader problem we are trying to answer is one that crosses various domains and
engineering fields. To put it quite simply, \textit{programs crash}. Dependent
typed languages, long a toy for theoretical computer scientists and
constructivist mathematicians, are increasingly becoming realistic tools to
write code with necessary guarantees of correctness. In other words, dependent
types are a solution in search of a problem. 

In all of these applications, I found something incredibly interesting that I'd
like to reflect on. I don't deny that the process of understanding all of these
papers at the undergraduate-level has not been an easy undertaking. This is
important because the underlying discussion question about this literature
review has been making dependent types accessible to a practical programmer.
However, while writing out inductive proofs in Idris or building a relational
algebra are difficult tasks, the result is surprisingly easy for a programmer to
use. 
\begin{enumerate}
    \item A programmer wanting to use well-typed units of measurement can simply
    import a library of SI units without having to worry about proving lemmas
    that demonstrate two derived units are the same. 
    \item A programmer wanting to build a parser can simply describe the data
    using a provided data description language. 
    \item A programmer wanting to use a type-safe relational database simply has
    to express schemas and then write expressions in our relational algebra
    detailing the input and output schema. 
    \item A programmer looking to implement complicated pattern-matching for a
    domain-specific language like Cryptol can easily implement that complex
    pattern-matching, as well as many other complicated pattern-matching
    techniques once splitView and its associated lemmas are implemented. 
    \item A programmer who wants to use the algebraic effects DSL can just
    describe the side effects present in their function without worrying about
    the complex underlying dependently-typed machinations. 
\end{enumerate}

In other words, at the current state, I find dependent types to still be too
complicated for the average programmer to understand. What people who understand
dependent types can do however, is work on dependently typed libraries where the
building blocks of what a programmer might use are dependently typed. A
programmer who uses a dependently-typed relational algebra needn't understand
the complex underlying machinations of building such a relational algebra. They
merely need to understand that writing schemas and expressions in the relational
algebra give way to type safety. In other words, the problem of getting every
programmer to love and write programs with dependent types is certainly
non-trivial, but providing dependently-typed libraries can be easily done. You
can't make a kid love vegetables, but nothing's stopping you from slipping a
little spinach into their dinner. 

Eventually however, the libraries that we've described will still require a
programmer to deal with increasingly complex error messages. Programmers using
dependently-typed libraries will eventually need to dig down and fight with the
compiler. However, dependently-typed libraries that are easy to use provide
entry points for programmers to become familiar with the safety provided by
dependent types. In other words, as work continues on mainstreaming dependent
types, programmers and academics with a deep understanding of dependent types
who wish to see its proliferation should work on providing well-written
libraries that encourage practical programmers to give dependent types a shot.

\clearpage

\section{Conclusions from Writing Palpatine}

The second portion of this senior thesis, the writing of Palpatine,
taught me a great deal as well. Some of the conclusions I've written below. 

\subsection{STV is an Algorithm in Need of a Declarative Specification}

I'm incredibly concerned by the fact that there does not currently exist an
official, Australian Electoral Commision sanctioned specification of STV.
Legislation serves as something close to a \textit{formal specification}, but as
there are gaps. For example, legislation currently has no specification for
tiebreaking behaviour \cite{stv_haskell}. Currently, Palpatine will choose the
leftmost, first occurring highest scored candidate and elect that one. This
decision is found in the function \texttt{getHighestIndex} for our
\texttt{Candidates}, and whether or not if it sees two candidates with the same
score, it changes the highest index to the second occuring one or not. If it
chooses the first, it favours candidates on the left. If not, it favours
candidates on the right. This is obviously unfair, but occurs within the
confines of a legislative system that does not account for the unlikely event of
a tie. 

Given that there is no formal AEC sanctioned specification, I had to go about
implementing Palpatine based on a high-level description of counting ballots
provided on their website. Furthermore, in addition to a formal specification of
the algorithm's behaviour, I could not find a formal, declarative specification
for the algorithm's intent. What does it mean for the STV algorithm to
``succeed''? Why is it optimal to eliminate and elect candidates in the order we
do? What is the definition of an optimally accurate election result and how does
STV reflect that? These questions blur the line between mathematics, computer
Science and policy, but they are important questions and must be addressed. 

\subsection{Totality Checkers Are Good.}

Guarantees of termination are good for the user. STV vote counting can take a
while. When running the code from Ghale, Gore and Pattinson, counting STV
ballots for an election of the Australian Capital Territory took 3 days
\cite{stv_haskell}. I'm uncertain of why that is the case, but it potentially
has to do with inefficiencies of the Coq-to-Haskell extraction process used in
the paper. More than once, I wondered if the program was not going to terminate.
Idris provides a built-in totality checker, which as specified above, means that
there are compile-time guarantees of termination. This means that if I use the
totality checker, my program \textit{is} a proof of termination. 

While significant, non-trivial barriers exist to implementing a totality checker
in Haskell, the experience I had with implementing an STV counter in Idris leads
me to believe that Haskell would benefit significantly from such a totality
checker. As such, I add my voice to encourage the people who are working towards
this noble goal. 

\subsection{Better Programming Languages Enable Safer Code.}

The constraints that type systems create force programmers to write better code.
For example, in the interest of satisfying totality, as described above, there
were many points during the writing of Palpatine where I was forced to employ
usage of \texttt{Maybe} to account for the possibility of the lack of a return
value. By forcing me towards complete case coverage, and by writing tighter type
constraints, I became a better programmer. Over time, I embraced the concept of
\textit{Type Driven Development}, introduced by Brady, and would write out my
types in advance, filling the code in with typed holes and then progressively
expanding out until I had code that satisfied my type \cite{tdd_book}. While in
the beginning, I found this an exercise in frustration, over time I found it to
be a powerful and enjoyable paradigm to work in. 

\subsection{Idris Has Slight Rough Edges}

While I greatly enjoyed working in Idris, it still does not enjoy certain
features that Haskell does. Unlike Haskell, Idris does not support the automatic
\textit{deriving} keyword which allows a programmer to easily extend a type to
support type classes. I had to define the implementations for \texttt{Eq, Ord,
Show} for \texttt{Candidate} and all subsequent custom data types by hand. It
was especially frustrating for \texttt{Judgment}, which I felt was a very
generic ADT that Haskell's \texttt{deriving} keyword would've handled just fine.

Idris currently has no standardized package manager and I ended up using a build
tool called Idrin. I would love for the community to standardize around a single
build tool like Idrin and work on building a package repository so that we can
easily share dependently-typed programs and libraries. 

At this point, Idris successfully parses small text files containing data that
conforms to my file format specification. However, reading large files (hundreds
of kilobytes) produces a segmentation fault that is a result of an error in the
\texttt{readFile} function in the Idris Prelude. As such, it is currently not
possible to run Palpatine on large election-scale text files. A bug report has
been opened with the developers of Idris \footnote{The bug report is available
here: https://github.com/idris-lang/Idris-dev/issues/4419}. 

\subsection{Verification is Hard}

I started this senior thesis with the intent of fully `verifying' the STV vote
counting algorithm. This past year has been a process of humbling myself and of
realizing that the problems that I'm trying to solve are incredibly difficult.
As I learned what \textit{verification} truly was, I realized that I could not
find a formal specification for STV that would be conducive to the process of
\textit{verification}. Indeed, even with a formal specification, the task of
verification can often be incredibly difficult. The creators of COMPCERT, a
verified compiler for a subset of the C language written in Coq, note that the
Coq verification portion of their compiler was six times larger than the
compiler implementation itself \cite{compcert}. These problems are difficult,
but it has been liberating to know that I'm not alone in finding them difficult.


In the end, I have produced a vote counter that merely verifies the cardinality
of two data structures. It does not verify that the outcome is correct, nor does
it verify in any way that we have even elected the correct number of candidates.
While from one perspective, this could be seen as frustrating, I actually found
it to be ultimately rewarding. I gained even more respect for the noble goal of
verification and a deeper understanding of what it means for code to be correct.
