\textit{This section is a summary of Edwin Brady's "Programming and Reasoning
with Algebraic Effects and Dependent Types" \cite{algebraic}}

In functional programming, we want to isolate side effects as much as possible
to keep our code clear. In Haskell and many other programming languages, side
effects like IO, state, random number generation, etc. are handled by monads
\cite{realworldhaskell}\footnote{This part of the literature assumes that you
are familiar with Monads, which is how the Haskell programming language handles
side effects. For more information, I highly recommend reading O'Sullivan et.
al's Real World Haskell or other widely available tutorials on Monads in
Haskell.}. If we want to use several Monads at once (our code requires
simultaneous handling of different side effects), we are often required to use
monad transformers. While this approach works for programs that require one or
two transformations between monads, as we bring in more and more side effects,
the number of transformation monads we need to write increases quite quickly. 

Work exists to sidestep the problem of handling increasingly complex monadic
transformations by encoding algebraic effects as a domain-specific language in a
dependently typed programming language \cite{algebraic}. We can start by
defining an \texttt{EFFECT} type, as seen in Figure~\ref{effects_def}. This
\texttt{EFFECT} data type enumerates the different kinds of side effects that an
Effects DSL could need to handle. In order for a function to use our effects
DSL, it will have to be of type \texttt{Eff}, where \texttt{Eff} is a data
declaration where the 'execution context' m (optionally a Monad) is specified, a
list of side effects, and the program's return type. For example, a function
that of the execution context \texttt{IO} that throws side effects, does work on
STDIO, and maintains an integer state will look something like the function
\texttt{example} in Figure~\ref{effects_def}. 

\begin{figure}[ht!!!!!!!!!!!]
    \caption{Definition of effect type}
    \label{effects_def}
    \begin{lstlisting}
        data EFFECT : Type where
            STATE     : Type -> EFFECT
            EXCEPTION : Type -> EFFECT
            FILEIO    : Type -> EFFECT
            STDIO     : EFFECT
            RND       : EFFECT
        
        data Eff : (m : Type -> Type) -> (es : List EFFECT) -> 
            (a : Type) -> Type
    
        example : Eff IO [EXCEPTION String, STDIO, STATE Int] ()
    \end{lstlisting}
\end{figure}

\texttt{example} demonstrates the power of the \texttt{Effects} DSL. Whereas if
I were using monadic transformers, I would have a type signature that strings
together monadic transformations in a complicated manner, here, I merely state
an execution context and all the other side effects that are present. There
would also be nested calls to \texttt{lift} that would be difficult to
understand. A programmer could be motivated to instead define a
function-specific monad that contains all the side-effect operations that the
function is required to use, to avoid the problem of transformation between
monads. The problem is, by defining such a blob monad, we have a single,
amorphous side effect, and monads that must be constantly defined on a
per-function basis. As \texttt{example} shows, a function written in our Effects
DSL only has to specify its execution context and other enumerated side effects.
You can then use all the effects necessary for your program, with no monadic
transformations required, as if you were writing a \texttt{blob} monad. Thus, we
have the composability of monadic transformations with the simplicity of writing
amorphous blob monads. 

\texttt{EFFECT} enumerates different side effects that we'd expect an effects
DSL to handle for us. We want to define an \texttt{Effect} type which we can
express our side effects in terms of, and also, a type class \texttt{Handler}
that allows us to run expressions in a certain computational context. The
definition of these two items are given in Figure~\ref{effects_def}. Notice how
\texttt{Effect} is parameterized by an input type, output type, and type of the
computation. Handler allows us to specify some computational context and then
run relevant effects. These definitions will become more clear as we implement
a \texttt{State} effect. 

\begin{figure}[ht!!!!]
    \caption{Declaring handlers for side effects. Taken from Brady's work.
    \protect\cite{algebraic}}
    \label{handlers}
    \begin{lstlisting}

        data State : Effect where
            Get : State a a a
            Put : b -> State a b ()

        get : Eff m [STATE x] x
        get = mkEffect Get

        put : x -> Eff m [STATE x] ()
        put val = mkEffect (Put val)

        Effect : Type
        Effect = (res : Type) -> (res' : Type) -> (t : Type) -> Type

        class Handler (e : Effect) (m : Type -> Type) where
            handle : res -> (eff : e res res' t) -> (res' -> t -> m a) -> m a
        
        instance Handler State m where
            handle st Get     k = k st st
            handle st (Put n) k = k n ()
        
        instance Handler StdIO IO where
            handle () (PutStr s) k = do putStrs; k () ()
            handle () GetStr     k = do x <- getLine; k () x
    \end{lstlisting}
\end{figure}

The example that we will be dealing with later uses the \texttt{STATE} effect,
so we can look at its implementation in Figure~\ref{handlers}. We want
\texttt{STATE} to have a \texttt{get} and {put} operation, of which, we expect
\texttt{get} to be able to be called an unlimited amount of times without any
impact on the type of state. We would like \texttt{put} to potentially mutate
the type of \texttt{STATE}. We therefore, define our two effects in the data
class, where Get does not modify the resource held in \texttt{State} whatsoever,
so its effect looks exactly paramaters are all the same type. \texttt{Put}'s
mutation takes in a value of type \texttt{b} and changes the type held by
\texttt{State}. As a put operation, we expect nothing to be returned, so the
type of the computation is \texttt{()}.

This entire example makes extensive use of dependent types. Note in the
declaration of \texttt{Handler}, \texttt{Handler}'s type class first
parameterizes over an \texttt{Effect}. The value of the effect is then the type
signature of the second parameter of the \texttt{handle} function. If we look at
the subsequent declarations of the instance \texttt{Handler State m}, note how
by specifying the \texttt{Effect} as \texttt{State}, we are able to pattern
match on the \texttt{Get} and \texttt{Put} effects in the declaration of
\texttt{Effect}. The entire \texttt{Effects} DSL hinges on this relationship
because if these constraints were not specified, nothing would stop a programmer
from defining other Effects inside of our declaration of the \texttt{State}
effect. We would be able to put something that should exist as a \texttt{Random}
effect inside of the declaration, and the compiler would not be able to check
that the value of the \texttt{e} effect matches accordingly. In other words,
without dependent types, the Effects DSL may as well just not exist, since its
goals of providing safe isolation of side effects just aren't possible without a
dependently typed programming language.

Functions with different side effects are invoked from an execution context
\texttt{Env}. The specifics of how execution contexts are defined are not
important, however, we can note that different effects require different
execution contexts. \texttt{STATE}, for example, is strictly locally confined
and thus its execution context does not matter. Meanwhile, \texttt{STDIO} does
IO operations, and therefore its execution context must be inside of the IO
Monad. We can specify a type class \texttt{Handler}. All effects should be
instances of this type class. \texttt{Handler} enforces that the environment we
invoke a function from is the correct one. In other words, if I attempted to
invoke a \texttt{runpure} function that had an \texttt{STDIO} side effect, the
program will not compile. 

We can now apply this small Effects DSL we have defined to work on some simple
programs where we need to maintain a side effect of some sort. I will provide an
example of a program where we tag each node of a binary tree with a unique ID. 

\begin{figure}[ht!!!!]
    \caption{Tagging a binary tree with integers. Taken from Brady's work.
    \protect\cite{algebraic}}
    \label{tag_def}
    \begin{lstlisting}
        -- Simple type def of binary tree in Idris
        data Tree a = Leaf
                    | Node (Tree a) a (Tree a)
        
        -- Takes in a tree and produces a tagged tree with
        -- State containing an integer passed inside of
        -- the function. 
        tag : Tree a -> Eff m [STATE Int] (Tree (Int, a))
        tag Leaf = return Leaf
        tag (Node l x r) = do
            l' <- tag l
            lbl <- get; put (lbl + 1)
            r' <- tag r
            return (Node l' (lbl, x) r')
        
        EffM :  (m   : Type -> Type) ->
                (es  : List EFFECT) ->
                (es' : List EFFECT) ->
                (a   : Type) -> Type
        
        run : Applicative m => 
                Env m es -> EffM m es es' a -> m a
        runPure : Env id es -> EffM id es es' a -> a

        tagFrom : Int -> Tree a -> Tree (Int, a)
        tagFrom x t = runPure [x] (tag t)
    \end{lstlisting}
\end{figure}

This tagging program is a simple function that recurses through a tree. The
function \texttt{get}s from and \texttt{put}s to a state that is kept alive for
the duration of the program. Since this side effect is 'created' at function
call-time and 'destroyed' on termination, it makes sense to say that this
function can be invoked as a pure function. Thus, we have a function with
internal side effects that have been clearly specified. We can see here how
specifying side effects as a list of effects rather than with monad transformers
means that we can easily add or remove side effects as required. Let's say I
wanted to add IO code into \texttt{tag}. When calling it, all I have to do is
call it in the context of \texttt{run}, specifying the execution context of IO.
Monadic transformers would require me to make a type signature that chains
together different monadic transformers. I'd then be required to convert data
between different monads in inconvenient ways that'll only increase in
complexity as I require more side effects.

Isolating side effects is a form of best practice and doing so is not easy.
While monads provide a way to handle side effects, programs multiple side
effects often require the use of monad transformers, making a programmer more
likely to take coarse-grained shortcuts which avoid many of the benefits of
monads in the first place. Building an \texttt{Effects} DSL allows the
programmer to specify effects of a program that executes inside a single
execution context. This allows for multiple monadic side effects to be
effectively juggled by a programmer. Such an \texttt{Effects} DSL is made
possible through a dependently typed programming language.